Modal type theory based on the intuitionistic epistemic logic

### 2018-03-25 MT/References.tex
modify

\documentclass[a4paper]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bussproofs}
\usepackage{mathtools}
\usepackage{verbatim}
\usepackage{dsfont}
\usepackage{mathabx}
\usepackage[all, 2cell]{xy}
\usepackage[all]{xy}
\newtheorem{theorem}{Theorem}
\newtheorem{prop}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{defin}{Definition}
\newtheorem{ex}{Example}
\newtheorem{col}{Corollary}
\newtheorem{con}{Consequence}
\usepackage{listings} 		% for source code
\date{}
\title{Modal type theory based on the intuitionistic epistemic logic}

\begin{document}
\maketitle

\begin{abstract}
Modal intuitionistic epistemic logic IEL$^{-}$ was proposed by S.Artemov and T. Protopopescu as the formal
foundation for the intuitionistic theory of knowledge. We construct a modal simply typed lambda-calculus
which is Curry-Howard isomorphic to IEL$^{-}$ as formal theory of calculations with applicative functors in
functional programming languages like Haskell or Idris.
\end{abstract}

  % This is a comment, not shown in final output.
  % The following shows typesetting  power of LaTeX:

  \section{Introduction}

  Modal intutionistic epistemic logic IEL was proposed by S. Artemov and T. Proropopescu \cite{Artemov}. IEL
  provides the epistimology and the theory of knowledge as based on BHK-semantics of intuitionistic logic.
  IEL$^{-}$  is a variant of IEL, that corresponds to intuitionistic belief.
  Informally, ${\bf K}A$ denotes that $A$ is verified intuitionistically.

  Intuitionistic epistemic logic IEL$^{-}$ is defined by following axioms and derivation rules:

  \begin{defin} Intuitionistic epistemic logic IEL:

    1) IPC axioms;

    2) ${\bf K}(A \to B) \to ({\bf K}A \to {\bf K}B)$ (normality);

    3) $A \to {\bf K}A$ (co-reflection);

    Rule: MP.

  \end{defin}

  V. Krupski and  A. Yatmanov provided the sequential calculus for IEL and proved that this calculus is
  PSPACE-complete \cite{Krupski}.

  Functional programming languages such as Haskell \cite{Haskell}, Idris \cite{Idris}, Purescript \cite{Purs}
  Elm \cite{Elm} or Scala \cite{Scala} have special type classes\footnote{Type class in Haskell is a general interface for
  special group of datatypes.} for calculations with container types like \verb"Functor" and
  \verb"Applicative" \footnote{Reader may read more about container types in the Haskell standard library documentation\cite{Base} or in the next one textbook
  \cite{Miran}}:

  \begin{lstlisting}[language=Haskell]
    class Functor f where
      fmap :: (a -> b) -> f a -> f b

    class Functor f => Applicative f where
      pure :: a -> f a
      (<*>) :: f (a -> b) -> f a -> f b
  \end{lstlisting}

  By \emph{container} (or \emph{computational context}) type we mean some type-operator $f$, where $f$ is a
  ``function'' from $*$ to $*$: type operator takes a simple type (ща kind $*$) and returns another
  simple type of kind \verb"*". For more detailed description of the type system with kinds used in
  Haskell see \cite{Morten}.

Applicative functor allows to generalize the action of a functor for functions with arbitrary number of arguments, for instance:
  \begin{lstlisting}[language=Haskell]
  liftA2 :: Applicative f => (a -> b -> c) -> f a -> f b -> f c
  liftA2 f x y = ((pure f) <*> x) <*> y
  \end{lstlisting}

  It's not difficult to see that modal axioms in $IEL^{-}$ and types of the methods of Applicative class in
  Haskell-like languages (which is described below) are syntactically similar and we are going to show that
  this coincidence has a non-trivial computational meaning.

  We investigate the relationship between intuitionistic epistemic logic IEL$^{-}$ and applicative programming with side-effects
  by constructing the type system (which is called $\lambda_{{\bf K}}$) which is Curry-Howard isomorphic to $IEL^{-}$.
  So we will consider ${\bf K}$-modality as an arbitrary applicative functor and we prove that obtained type system
  is sound and complete for applicative functor on cartesian closed category (using the categorical definition proposed by Paterson \cite{Cons}).

  $\lambda_{{\bf K}}$ consists of the rules for simply typed lambda-calculus and special typing rules for
  lifting types into the applicative functor ${\bf K}$. We assume that our type system will
  axiomatize the simplest case of computation with effects with one container. We provide a proof-theoretical
  view at this kind of computations in functional programming and prove strong normalization and confluence.

  \section{Typed lambda-calculus based on IEL$^{-}$}

  The first is to define the natural deduction calculus for NIEL$^{-}$ :

  \begin{defin} Natural deduction NIEL$^{-}$ for IEL$^{-}$ is an extenstion of intuitionistic natural deduction calculus
    with additional inference rules for modality:

  \begin{minipage}{0.5\textwidth}
    \begin{flushleft}
  	\begin{prooftree}
    	\AxiomC{$\Gamma \vdash A$}
    	\RightLabel{${\bf K}_I$}
    	\UnaryInfC{$\Gamma \vdash {\bf K}A$}
  \end{prooftree}
    \end{flushleft}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \begin{flushright}
  	\begin{prooftree}
    \AxiomC{$\Gamma \vdash {\bf K} A_1, \dots, \Gamma \vdash {\bf K} A_n $}
    \AxiomC{$A_1,\dots,A_n \vdash B$}
    \BinaryInfC{$\Gamma \vdash {\bf K} B$}
  	\end{prooftree}
    \end{flushright}
  \end{minipage}
  \end{defin}

The first rule allows to derive co-reflextion. The second modal rule is a counterpart of $\Box_I$
rule in natural deduction calculus for constructive K (see \cite{ModalLa}).

We will denote $\Gamma \vdash {\bf K} A_1, \dots, \Gamma \vdash {\bf K} A_n$ and $A_1,\dots,A_n \vdash B$ as $\Gamma \vdash {\bf K} \vec{A}$ and $\vec{A} \vdash B$ for brevity.

  \vspace{\baselineskip}

  \begin{lemma}
    $\Gamma \vdash_{\text{NIEL}^{-}} A \Rightarrow$ IEL$^{-} \vdash \bigwedge \Gamma \rightarrow A$.
  \end{lemma}

  \begin{proof}
Induction on the derivation.

  \vspace{\baselineskip}

Let us consider cases with modality.

\vspace{\baselineskip}

1) If $\Gamma \vdash_{\text{NIEL}^{-}} A$, then $\text{IEL}^{-} \vdash \bigwedge \Gamma \rightarrow {\bf K}A$.

$\begin{array}{lll}
(1) & \bigwedge \Gamma \rightarrow A & \text{assumption}\\
(2) & A \rightarrow {\bf K}A &\text{co-reflection}\\
(3) & (\bigwedge \Gamma \rightarrow A) \rightarrow ((A \rightarrow {\bf K}A) \rightarrow (\bigwedge \Gamma \rightarrow {\bf K}A))&\text{IPC theorem}\\
(4) & (A \rightarrow {\bf K}A) \rightarrow (\bigwedge \Gamma \rightarrow {\bf K}A) &\text{from (1), (3) and MP}\\
(5) & \bigwedge \Gamma \rightarrow {\bf K}A &\text{from (2), (4) and MP}\\
\end{array}$

\vspace{\baselineskip}

2) If $\Gamma \vdash_{\text{NIEL}^{-}} {\bf K} \vec{A}$ and $\vec{A} \vdash B$, then $\text{IEL}^{-} \vdash \bigwedge \Gamma \rightarrow {\bf K}B$.

$\begin{array}{lll}
(1) &\bigwedge \Gamma \rightarrow \bigwedge \limits_{i = 1}^{n} {\bf K} A_i & \text{assumption} \\
(2) &\bigwedge \limits_{i = 1}^{n} {\bf K} A_i \rightarrow {\bf K} \bigwedge \limits_{i = 1}^{n} A_i& \text{IEL$^{-}$ theorem} \\
(3) &\bigwedge \Gamma \rightarrow {\bf K} \bigwedge \limits_{i = 1}^{n} A_i & \text{from (1), (2) and transitivity} \\
(4) &\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B& \text{assumption} \\
(5) &(\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B) \rightarrow {\bf K} (\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B)& \text{co-reflection}\\
(6) &{\bf K} (\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B)& \text{from (4), (5) and MP} \\
(7) &{\bf K} \bigwedge \limits_{i = 1}^{n} A_i \rightarrow {\bf K} B & \text{from (6) and normality} \\
(8) &\bigwedge \Gamma \rightarrow {\bf K} B & \text{from (3), (7) and transitivity}
\end{array}$

  \end{proof}

\begin{lemma}
  $ $
  If $\text{IEL}^{-} \vdash A$, then $\text{NIEL}^{-} \vdash A$.
\end{lemma}

\begin{proof}
  Straightforward derivation of modal axioms in $\text{NIEL}^{-}$. We consider this derivation below using terms.
\end{proof}

  \vspace{\baselineskip}

  At the next step we build the typed lambda-calculus based on the NIEL$^{-}$ with implication and dijunction
by proof-assingment in rules. Obtained fragment is equivalent to IEL$^{-}$ without axioms for negation and disjunction.

  At first, we define lambda-terms and types for this lambda-calculus.

  \vspace{\baselineskip}

  \begin{defin} The set of terms:

  Let $\mathbb{V}$ be the set of variables. The set $\Lambda_{{\bf K}}$ of terms is defined by the grammar:

$\begin{array}{lll}
& \Lambda_{{\bf K}} ::= \mathbb{V} \: | \:  (\lambda \mathbb{V}.\Lambda_{{\bf K}}) \: | \: (\Lambda_{{\bf K}}\Lambda_{{\bf K}}) \: | \: (\Lambda_{{\bf K}} , \Lambda_{{\bf K}}) \: | \: (\pi_1 \Lambda_{{\bf K}}) \: | \: (\pi_2 \Lambda_{{\bf K}}) \: | & \\
& \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad ({\bf pure \: } \: \Lambda_{{\bf K}}) \: | \: ({\bf let \:\: pure} \:\: \mathbb{V}^{*} = \Lambda_{{\bf K}}^{*} \:\: {\bf in} \:\: \Lambda_{{\bf K}})
\end{array}$

  \end{defin}

Where $\mathbb{V}^{*}$ and $\Lambda_{{\bf K}}^{*}$ denote the set of finite sequences of variables $\bigcup \limits_{i=0}^{\infty} \mathbb{V}^i$
and the set of finite sequences of terms $\bigcup \limits_{i = 0}^{\infty} {\Lambda_{{\bf K}}}^i $. Note that the sequence of variables $\vec{x}$ and the sequence of terms $\vec{M}$ should have the same length.
Otherwise, term is not well-formed.

  \begin{defin} The set of types:

  Let $\mathbb{T}$ be the set of atomic types. The set $\mathbb{T}_{{\bf K}}$ of types with applicative
  functor ${\bf K}$ is generated by the grammar:
  \begin{equation}
    \mathbb{T}_{{\bf K}} ::= \mathbb{T} \: | \: (\mathbb{T}_{{\bf K}} \to \mathbb{T}_{{\bf K}}) \: |
    \: (\mathbb{T}_{{\bf K}} \times \mathbb{T}_{{\bf K}}) \: | \: ({\bf K}\mathbb{T}_{{\bf K}})
  \end{equation}
  \end{defin}

  Context, domain of context and range of context are defined standardly \cite{Neder}\cite{Morten}.

  Our type system is based on the Curry-style typing rules:

  \begin{defin} Modal typed lambda calculus $\lambda_{{\bf K}}$ based on NIEL$^{-}_{\land, \to}$:

    \begin{center}
    \begin{prooftree}
    \AxiomC{$ $}
    \RightLabel{\scriptsize{ax}}
    \UnaryInfC{$\Gamma , x : A \vdash x : A$}
    \end{prooftree}
    \end{center}

    \begin{minipage}{0.45\textwidth}
      \begin{prooftree}
      \AxiomC{$\Gamma, x : A \vdash M : B$}
      \RightLabel{$\rightarrow_i$}
      \UnaryInfC{$\Gamma \vdash \lambda x. M : A \to B$}
      \end{prooftree}

      \begin{prooftree}
      \AxiomC{ $\Gamma \vdash M : A$ }
      \AxiomC{ $\Gamma \vdash N : B$ }
      \RightLabel{$\times_i$}
      \BinaryInfC{$\Gamma \vdash \langle M, N \rangle : A \times B$}
      \end{prooftree}

      \begin{prooftree}
        \AxiomC{$\Gamma \vdash M : A$}
        \RightLabel{${\bf K}_I$}
        \UnaryInfC{$\Gamma \vdash {\bf pure \: } \: M : {\bf K}A $}
      \end{prooftree}
  \end{minipage}%
  \hfill
  \begin{minipage}{0.45\textwidth}
  \begin{tabular}{p{\textwidth}}
    \begin{prooftree}
    \AxiomC{$\Gamma \vdash M : A \to B$}
    \AxiomC{$\Gamma \vdash N : A$}
    \RightLabel{$\rightarrow_e$}
    \BinaryInfC{$\Gamma \vdash MN : B$}
    \end{prooftree}

    \begin{prooftree}
    \AxiomC{ $\Gamma \vdash M : A_1 \times A_2$ }
    \RightLabel{$\times_e$, $i \in \{ 1, 2 \}$}
    \UnaryInfC{$\Gamma \vdash \pi_i M : A_i$}
    \end{prooftree}

    \begin{prooftree}
      \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
      \AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
      \RightLabel{$\text{let}_{{\bf K}}$}
      \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N : {\bf K} B$}
    \end{prooftree}
  \end{tabular}
  \end{minipage}%

  \end{defin}

  ${\bf K}_I$-typing rule is the same as $\bigcirc$-introduction in lax logic (also known as monadic
  metalanguage \cite{Lax}) and in typed lambda-calculus which is derived by proof-assignment for lax-logic
  proofs.
  ${\bf K}_I$ allows to inject an object of type $\alpha$ into the functor. ${\bf K}_I$  reflects the
  Haskell method {\bf pure} for Applicative class. It plays the same role as the {\bf return} method in Monad class.

  $\text{let}_{{\bf K}}$ is similar to the $\Box$-rule in typed lambda calculus for intuitionistic normal modal logic {\bf IK}, which is described in \cite{ModalK}.

$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ is a syntax sugar for the sequence $\Gamma \vdash M_1 : {\bf K}A_1,\dots,\Gamma \vdash M_n : {\bf K}A_n$ and $\vec{x} : \vec{A} \vdash N : B$ is a short form for $x_1 : A_1, \dots, x_n : A_n \vdash N : B$.
${\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N$ is a simultaneous local binding in $N$.
We use this short form instead of ${\bf let \: pure\:} x_1,\dots,x_n = M_1,\dots,M_n {\: \bf in \:} N$.

In fact, our calculus is the extention of typed lambda calculus for {\bf IK} by ${\bf K}_I$-rule that is appropriate to co-reflection.

  \vspace{\baselineskip}

Here are some examples of closed terms:

\begin{prooftree}
\AxiomC{$x : A \vdash x : A$}
\UnaryInfC{$x : A \vdash {\bf pure \:} x : {\bf K} A$}
\UnaryInfC{$\vdash (\lambda x. {\bf pure \: } x) : A \to {\bf K}A$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$f : {\bf K} (A \to B) \vdash f : {\bf K} (A \to B)$}
\AxiomC{$x : {\bf K} A \vdash x : {\bf K} A $}
\AxiomC{$g : A \to B \vdash g : A \to B$}
\AxiomC{$y : A \vdash y : A$}
\RightLabel{$\to_e$}
\BinaryInfC{$g : A \to B, y : A \vdash g y : B$}
\RightLabel{$\text{let}_{{\bf K}}$}
\TrinaryInfC{$f : {\bf K} (A \to B), x : {\bf K} A \vdash {\bf let \: pure \:} g, y  = f, x {\: \bf in \:} g y : {\bf K} B$}
\UnaryInfC{$f : {\bf K} (A \to B) \vdash \lambda x. {\bf let \: pure\:} g, y = f, x {\: \bf in \:} g y : {\bf K} A \to {\bf K} B$}
\UnaryInfC{$\vdash \lambda f. \lambda x. {\bf let \:pure \:} g, y = f, x {\: \bf in \:} g y : {\bf K}(A \to B) \to {\bf K} A \to {\bf K} B$}
\end{prooftree}

  \vspace{\baselineskip}

  Now we define free variables and substitutions. $\beta$-reduction, multi-step $\beta$-reduction and $\beta$-equality are defined standardly:

  \begin{defin} The set $FV(M)$ of free variables for a term $M$:

  1) $FV(x) = \{ x \}$;

  2) $FV(\lambda x. M) = FV(M) \setminus \{ x\}$;

  3) $FV(M N) = FV(M) \cup FV(N);$

  4) $FV(\langle M,N \rangle) = FV(M) \cup FV(N)$;

  5) $FV(\pi_i M) \subseteq FV(M)$, $i \in \{ 1, 2\}$;

  6) $FV(\text{pure } M) = FV(M)$;

  7) $FV({\bf let \: pure} \: \vec{x} = \vec{M} \:\: {\bf in} \:\: N) = \bigcup \limits_{i = 1}^n FV(M), \text{where $n = |\vec{M}|$}$.
  \end{defin}

  \begin{defin} Substitution:

  1) $x [x := N] = N$, $x [y := N] = x$;

  2) $(M N) [x := N] = M[x := N] N [x := N]$;

  3) $(\lambda x. M) [x := N] = \lambda x. M [y := N]$, $y \in FV(M)$;

  4) $(M, N)[x := P] = (M[x := P], N [x := P])$;

  5) $(\pi_i M) [x := P] = \pi_i (M[x := P])$, $i \in \{ 1, 2\}$;

  6) $({\bf pure \: } M) [x := P] = {\bf pure \: } (M [x := P])$;

  7) $({\bf let \: pure \:}\vec{x} = \vec{M} {\: \bf in \:} N) [y := P] = {\bf let \: pure \:} \vec{x} = (\vec{M} [y := P]) {\: \bf in \:} N$.
  \end{defin}

Substitution and free variable for terms of kind ${\bf let \: pure}$ are defined similary to \cite{ModalK}.

  \begin{defin} Type substituition

    The substituition of type $C$ for type variable $B$ in type $A$ is defined inductively:

    1) $B [B := C] = B$ and $D [B := C] = D$, if $B \neq D$;

    2) $(A_1 \alpha A_2)[B := C] = (A_1[B := C]) \alpha (A_2[B := C])$, where $\alpha \in \{ \to, \times \}$;

    3) $({\bf K}A)[B := C] = {\bf K}(A [B := C])$;

    4) Let $\Gamma$ be the context, then $\Gamma [B := C] = \{ x : (A[B := C]) \: | \: x : A \in \Gamma \}$.
  \end{defin}

  \begin{defin} $\beta$-reduction and $\eta$-reduction rules for $\lambda_{{\bf K}}$.

  1) $(\lambda x. M) N \rightarrow_{\beta} M [x := N]$;

  2) $\pi_1 \langle M, N \rangle \rightarrow_{\beta} M$;

  3) $\pi_2 \langle M, N \rangle \rightarrow_{\beta} N$;

  4) $\begin{array}{llll}
  & {\bf let \: pure \:} \vec{x}, y, \vec{z} = \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\: \bf in \: } Q, \vec{P} {\: in \:} R \rightarrow_{\beta} & \\
  & {\bf let \: pure \:} \vec{x}, \vec{w}, \vec{z} = \vec{M}, \vec{N}, \vec{P} {\: \bf in \: } R [y := Q]
  \end{array}$

  5) ${\bf let \: pure \:} \vec{x} = {\bf pure \:} \vec{M} {\: \bf in \:} N \rightarrow_{\beta} {\bf pure \:} N [\vec{x} := \vec{M}]$

  6) ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} M \rightarrow_{\beta} {\bf pure \:} M$, where \underline{\quad} is an empty sequence of terms.

  7) $\lambda x. f x \rightarrow_{\eta} f$;

  8) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

  9) ${\bf let \: pure \:} x = M {\: \bf in \: } x \rightarrow_{\eta} M$;

  \end{defin}

By default we use call-by-name evaluation strategy.

Now we will prove standard lemmas for contexts in type systems\footnote{We will not prove cases with
$\to$-constructor, they are proved standardly in the same lemmas for simply typed lambda calculus, for
example see \cite{Neder} \cite{Morten}. We will consider only modal cases}:

\begin{lemma} Generation for ${\bf K}_I$.

  Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$, then $\Gamma \vdash M : A$;
\end{lemma}

\begin{proof}
  Straightforwardly.
\end{proof}

\begin{lemma} Basic lemmas.

\begin{itemize}
  \item If $\Gamma \vdash M : A$ and $\Gamma \subseteq \Delta$, then $\Delta \vdash M : A$;
  \item If $\Gamma \vdash M : A$, then $\Delta \vdash M : A$, where $\Delta = \{ x_i : A_i \: | \: (x_i : A_i) \in \Gamma \: \& \: x_i \in FV(M) \}$
  \item If $\Gamma, x : A \vdash M : B$ and $\Gamma \vdash N : A$, then $\Gamma \vdash M [x := N] : B$.
  \item If $\Gamma \vdash M : A$, then $\Gamma [B := C] \vdash M : (A [B := C])$.
\end{itemize}
\end{lemma}

\begin{proof}
  $ $

1) The derivation ends in

\begin{prooftree}
  \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
  \AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
  \RightLabel{$\text{let}_{{\bf K}}$}
  \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N : {\bf K}B$}
\end{prooftree}

By IH $\Delta \vdash \vec{M} : {\bf K} \vec{A}$, so $\Delta \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N : {\bf K}B$.

  \vspace{\baselineskip}

2) Similary.

  \vspace{\baselineskip}

3) The derivation ends in

\begin{prooftree}
  \AxiomC{$\Gamma, z : C \vdash \vec{M} : {\bf K} \vec{A}$}
  \AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
  \RightLabel{$\text{let}_{{\bf K}}$}
  \BinaryInfC{$\Gamma, z : C \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N : {\bf K}B$}
\end{prooftree}

Let $\Gamma \vdash P : C$.

By IH, $\Gamma \vdash \vec{M} [z := P] : {\bf K} \vec{A}$. So $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} [z := P] {\: \bf in \: } N : {\bf K}B$

  \vspace{\baselineskip}

4-5) Similary.

\end{proof}

\begin{theorem} Subject reduction

  If $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\beta \eta} N$, then $\Gamma \vdash N : A$

\end{theorem}

\begin{proof}

Induction on the derivation $\Gamma \vdash M : A$ and on the generation of $\rightarrow_{\beta \eta}$.

For cases with application, abstraction and pairs see \cite{Morten} \cite{Pierce}.

1) If $\Gamma \vdash {\bf let \: pure \:} \vec{x}, y, \vec{z} = \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P}  {\: \bf in \:} R : {\bf K}B$,
  then $\Gamma \vdash {\bf let \: pure \:} \vec{x}, \vec{w}, \vec{z} = \vec{M}, \vec{N}, \vec{P} \: { \bf in } \: R [y := Q] : {\bf K}B$ by rule 4).

2) Let $\Gamma \vdash {\bf let \: pure \:} x = M {\: \bf in \: } x : {\bf K}A$, then $\Gamma \vdash M : {\bf K}A$ by rule 9).

See \cite{ModalK}.

3) The derivation ends in

\begin{prooftree}
\AxiomC{$\Gamma \vdash {\bf pure \:} \vec{M} : {\bf K}\vec{A}$}
\AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
\BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = {\bf pure \:} \vec{M} {\: \bf in \:} N : {\bf K}B$}
\end{prooftree}

So $\Gamma \vdash \vec{M} : \vec{A}$ by Lemma 4 and $\Gamma \vdash N [\vec{x} := \vec{M}] : B$ by Lemma 4, part 3.

Then we can transform this into the following derivation:

\begin{prooftree}
\AxiomC{$\Gamma \vdash N [\vec{x} := \vec{M}] : B$}
\RightLabel{${\bf K}_I$}
\UnaryInfC{$\Gamma \vdash {\bf pure \:} N [\vec{x} := \vec{M}] : {\bf K}B$}
\end{prooftree}

4) The derivation ends in

\begin{prooftree}
\AxiomC{$\vdash M : A$}
\UnaryInfC{$\vdash {\bf let \: pure \:} \underline{\quad} = \underline {\quad} {\: \bf in \:} M : {\bf K} A$}
\end{prooftree}

So, if $\vdash M : A$, then $\vdash {\bf pure \:} M : {\bf K}A$.

Note that this part of the lemma works conversly too.
\end{proof}

\begin{theorem}
  $ $

  $\twoheadrightarrow_{\beta}$ is strongly normalizing;
\end{theorem}

\begin{proof}
  $ $

We modify and apply Tait's technique of logical relation for modalities. See \cite{Pierce} \cite{Troelstra}.

We treat only modal cases below.

\begin{defin} The set of strongly computable terms:
  \begin{itemize}
    \item $SC_A = \{ M : A \: | \: M \text{ is strongly normalizing} \}$ for $A \in \mathbb{T}$;
    \item $SC_{A \to B} = \{ M : A \to B \: | \: \forall N \in SC_A, M N \in SC_B \}$, for $A,B \in \mathbb{T}_{\bf K}$ for $A, B \in \mathbb{T}_{{\bf K}}$;
    \item $SC_{{\bf K}A} = \{ M : {\bf K}A \: | \: M \text{ is strongly normalizing} \}$ for $A \in \mathbb{T}$;
    \item $\forall i \in \{ 1,\dots,n\}, \prod \limits_{i = 1}^{n} SC_{{\bf K}A_i} = \{ \vec{M} = (M_1,\dots,M_n) \:\: | \:\: \forall N \in SC_{B}, FV(N) = \{ x_1,\dots,x_n\} \: \& \: \forall i, x_i \in SC_{A_i} \Rightarrow {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N \in SC_{{\bf K} B}\}$
  \end{itemize}
\end{defin}

\begin{defin}
  A term $M$ is neutral, if it has one of the following forms:
  \begin{itemize}
    \item $M N$;
    \item If $M$ is neutral, then ${\bf pure \:} M$ is neutral;
    \item If $\vec{M}$ and $N$ are neutral, then ${\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: N}$ is neutral.
    $\vec{x}$ is a sequence of free variables of a term $N$.
  \end{itemize}
\end{defin}

\begin{lemma}
  $ $

\begin{itemize}
\item If $M \in SC_A$ and $A \in \mathbb{T}_{{\bf K}}$, then $M$ is strongly normalizing;
\item If $M \in SC_A$, $A \in \mathbb{T}_{{\bf K}}$ and $M \rightarrow_{\beta} N$, then $N \in SC_A$;
\item If $N$ is neutral, $N \in SC_A$. Then, if $M \rightarrow_{\beta} N$, then $M \in SC_A$;
\end{itemize}
\end{lemma}

\begin{proof}
  $ $

  By induction on the structure of $A$.

1) $A \equiv {\bf K}A$, where $A \in \mathbb{T}$.

i-ii-iii) Immediately.

    \vspace{\baselineskip}

2)

i) Suppose $\vec{M} = (M_1,\dots,M_n) \in \prod \limits_{i = 1}^{n} SC_{{\bf K}A_i}$.

Let $N \in SC_{B}$, such that $FV(N) = \{ x_1,\dots, x_n \}$ and $\forall i, x_i \in SC_{A_i}$.

So ${\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N \in SC_{{\bf K}B}$ by IH.

So $\vec{M}$ are strongly normalizing, since ${\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } N$ is strongly normalizing by IH.

\vspace{\baselineskip}

ii) Let $\vec{M_1} \in \prod \limits_{i = 1}^{n} SC_{{\bf K}A_i}$ and $\vec{M_1} \rightarrow_{\beta} \vec{M_2}$.

Let $N \in SC_{B}$, such that $FV(N) = \{ x_1,\dots,x_n \}$ and $\forall i, x_i \in SC_{A_i}$.

So ${\bf let \: pure \:} \vec{x} = \vec{M_1} {\: \bf in \:} N \rightarrow_{\beta} {\bf let \: pure \:} \vec{x} = \vec{M_2} {\: \bf in \:} N$

and ${\bf let \: pure \:} \vec{x} = \vec{M_2} {\: \bf in \:} N \in SC_{{\bf K}B}$ by assumption.

So $\vec{M_2} \in \prod \limits_{i = 1}^{n} SC_{{\bf K}A_i}$.

iii) Let $M_2$ be neutral, $M_2 \in \prod \limits_{i = 1}^{n} SC_{{\bf K}A_i}$ and $M_1 \rightarrow_{\beta} M_2$.

Let $N \in SC_{B}$, , such that $FV(N) = \{ x_1,\dots,x_n \}$ and $\forall i, x_i \in SC_{A_i}$.

So ${\bf let \: pure \:} \vec{x} = \vec{M_2} {\: \bf in \:} N \in SC_{{\bf K}B}$.

Thus ${\bf let \: pure \:} \vec{x} = \vec{M_1} {\: \bf in \:} N \rightarrow_{\beta} {\bf let \: pure \:} \vec{x} = \vec{M_2} {\: \bf in \:} \in N$.

Hence ${\bf let \: pure \:} \vec{x} = \vec{M_1} {\: \bf in \:} N\in SC_{{\bf K}B}$ by IH, so $\vec{M_1} \in \prod \limits_{i = 1}^{n} SC_{{\bf K}A_i}$.
\end{proof}

\begin{lemma}
  $ $

  If $M \in SC_A$, then ${\bf pure \:} M \in SC_{{\bf K}A}$
\end{lemma}

\begin{proof}

  Induction on the structure of M.
\end{proof}

\begin{lemma}
  $ $

  Let $x_1 : A_1,\dots, x_n : A_n \vdash M : A$, then forall $i, M_i \in SC_{A_i}$. Then $M [x_1 := M_1,\dots,x_n := M_n] \in SC_A$.
\end{lemma}

\begin{proof}
  $ $

Induction on the derivation of $x_1 : A_1,\dots, x_n : A_n \vdash M : A$.

  1) The derivation ends in:

\begin{prooftree}
\AxiomC{$x_1 : A_1,\dots, x_n : A_n \vdash M : A$}
\UnaryInfC{$x_1 : A_1,\dots, x_n : A_n \vdash {\bf pure \:} M : {\bf K}A$}
\end{prooftree}

  By assumption $M [x_1 := M_1,\dots,x_n := M_n] \in SC_A$, so ${\bf pure \:} M [x_1 := M_1,\dots,x_n := M_n] \in SC_{{\bf K}A}$.

\vspace{\baselineskip}

  2) The derivation ends in:
\begin{prooftree}
  \AxiomC{$x_1 : A_1,\dots, x_n : A_n \vdash \vec{M^{'}} : {\bf K}\vec{A}$}
  \AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
  \BinaryInfC{$x_1 : A_1,\dots, x_n : A_n \vdash {\bf let \: pure \:} \vec{x} = \vec{M^{'}} {\: \bf in \:} N : {\bf K}B$}
\end{prooftree}

By IH forall $i \in \{ 1,\dots,\text{length}(\vec{M^{'}}) \}$, $M_i^{'} [x_1 := M_1,\dots,x_n := M_n] \in SC_{{\bf K}A_i}$.

So ${\bf let \: pure \:} \vec{x} = \vec{M^{'}} [x_1 := M_1,\dots,x_n := M_n] {\: \bf in \:} N \in SC_{{\bf K} B}$,
otherwise we can build infinite reduction path in $\vec{M^{'}} [x_1 := M_1,\dots,x_n := M_n]$.
\end{proof}

\begin{col}

  All terms are strongly computable, therefore are strongly normalizing.
\end{col}

\end{proof}

\begin{theorem}
  $ $

  $\twoheadrightarrow_{\beta}$ is confluent.
\end{theorem}

\begin{proof}

  We modify and apply Barendregt's technique with term underlying. We will consider the grammar without constructors for pairs.

\begin{defin} The set of underlined terms.

  \begin{itemize}
    \item $x \in \mathbb{V} \Rightarrow x \in \underline{\Lambda}$;
    \item $M \in \underline{\Lambda} \Rightarrow (\lambda x. M) \in \underline{\Lambda}$;
    \item $M, N \in \underline{\Lambda} \Rightarrow (M N) \in \underline{\Lambda}$;
    \item $M \in \underline{\Lambda} \Rightarrow ({\bf pure \:} M) \in \underline{\Lambda}$;
    \item $\vec{x} \in \mathbb{V}^{*}, \vec{M} \in \underline{\Lambda}^{*}, N \in \underline{\Lambda} \Rightarrow {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N \in \underline{\Lambda}$, where $length(\vec{x}) = length(\vec{M})$;
    \item $M, N \in \underline{\Lambda} \Rightarrow (\lambda_i x. M) N \in \underline{\Lambda}$, for all $i \in \mathbb{N}$.
  \end{itemize}
\end{defin}

\begin{defin} Subsitution for term with labelled lambda:

  $((\lambda_i x. M) N) [y := Z] = (\lambda_i x. M [y := Z]) (N [y := Z])$
\end{defin}

\begin{defin} Index erasing

  Let us define map $|.| : \underline{\Lambda} \to \Lambda$ as follows:

>>>>

\addcontentsline{toc}{section}{Список использованной литературы}
